---
title: "Influence Functions"
author: "Cyrus Samii"
date: "3/13/2019"
output: 
  html_document:
    number_sections: yes
    df_print: paged
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Influence functions play an important role in semi-parametric theory. Informally, the influence function measures the influence of a single observation ona parameter estimate.  For good introductions, see Kennedy (2016, [arxiv](https://arxiv.org/abs/1510.04740)) and Kennedy (2017, [arxiv](https://arxiv.org/abs/1709.06418)).  Here I derive some influence functions for common estimators just to illustrate how the concept applies.

# Influence function of OLS

## Simulated data for the illustrations
```{r}
# Make data
n <- 16
x1 <- rnorm(n)
x2 <- rnorm(n)
e <- rnorm(n)
y <- x1 + x2 + e
X <- cbind(rep(1,n),x1,x2)

# Functions for summation representation
crossProdxx <- function(x){
	as.matrix(unlist(x))%*%t(as.matrix(unlist(x)))
}

crossProdxy <- function(x){
	xv <- unlist(x)
	l.x <- length(xv)
	as.matrix(xv[1:(l.x-1)])*xv[l.x]
}
```

## Two representations of OLS

We have matrix representation,
$$
\widehat{\beta} = (\mathbf{X}'\mathbf{X})^(-1)\mathbf{X}'Y
$$
as well as summation representation,
$$
\widehat{\beta} = \left(\sum_{i=1}^N X_iX_i' \right)^{-1}\sum_{i=1}^NX_iY_i
$$
```{r}
# Matrix representation:
solve(t(X)%*%X)%*%t(X)%*%y

# Summation representations
XX <- lapply(apply(X, 1, function(x){list(x)}),
		crossProdxx)
Xy <- lapply(apply(cbind(X,y), 1, function(x){list(x)}),
		crossProdxy)
Xe <- lapply(apply(cbind(X,e), 1, function(x){list(x)}),
		crossProdxy)
solve(Reduce("+", XX))%*%Reduce("+", Xy)
```

We can also write in terms of true $\beta$,
$$
\widehat{\beta} = \beta + \left(\sum_{i=1}^N X_iX_i' \right)^{-1}\sum_{i=1}^NX_ie_i,
$$
where $e_i$ is the population residual:
```{r}
as.matrix(c(0,1,1)) + solve(Reduce("+", XX))%*%Reduce("+", Xe)
```

## Deriving the influence function

We work with
\begin{align*}
\sqrt{N}(\widehat{\beta} - \beta) & = \left(\frac{1}{N}\sum_{i=1}^N X_iX_i' \right)^{-1}\frac{1}{\sqrt{N}}\sum_{i=1}^NX_ie_i \\
& = \frac{1}{\sqrt{N}}\sum_{i=1}^N \left(\frac{1}{N}\sum_{i=1}^N X_iX_i' \right)^{-1}X_ie_i \\
& = \frac{1}{\sqrt{N}}\sum_{i=1}^N E[X_iX_i]^{-1}X_ie_i + o_N(1).
\end{align*}


Equivalence of the first and second lines:

```{r}
solve((1/n)*Reduce("+", XX))%*%((1/sqrt(n))*Reduce("+", Xe))

(1/sqrt(n))*Reduce("+",lapply(Xe,
		function(x){solve((1/n)*Reduce("+", XX))%*%x}))
```

So we have that the OLS influence function is given by,
$$
\phi_{OLS}(X) = E[XX]^{-1}Xe
$$

We can compute the empirical analogues of the influence function values for each observation:
```{r}
lapply(Xe,
		function(x){solve((1/n)*Reduce("+", XX))%*%x})
```

OLS is a linear regular estimator, and so it abides by
$$
\sqrt{N}(\widehat{\beta} - \beta) \overset{p}{\rightarrow} \text{MVN}\left(\mathbf{0}, E[\phi_{OLS}(X)\phi_{OLS}(X)'] \right) = \text{MVN}\left(\mathbf{0}, E[XX]^{-1}E[XX'e^2]E[XX]^{-1} \right) .
$$

# Influence Function of IPW

For a binary treatment variable, $A$, consider the Horvitz-Thompson type IPW estimator,
$$
\widehat{\psi} = \frac{1}{N}\sum_{i=1}^N\left(\frac{A_iY_i}{\pi(X_i)} - \frac{(1-A_i)Y_i}{1-\pi(X_i)} \right),
$$

where $\pi(\cdot)$ is the propensity score function.

Now,
$$
\sqrt{N}(\widehat{\psi} - \psi) = \frac{1}{\sqrt{N}}\sum_{i=1}^N\left(\frac{A_iY_i}{\pi(X_i)} - \frac{(1-A_i)Y_i}{1-\pi(X_i)}  - \psi\right),
$$
which shows that the influence function for IPW is simply,
$$
\phi_{IPW} = \frac{AY}{\pi(X)} - \frac{(1-A)Y}{1-\pi(X)} - \psi.
$$
